---
title: "Assignment 1- EDA and Data Preprocessing on Kenyan Student
Dataset"
output: pdf_document
date: "2025-06-12"
---
```{r Install required libraries}
install.packages("tidyverse")
install.packages("corrplot")
install.packages("ggplot2")
install.packages("dplyr")

#Load the libraries
library(tidyverse)
library(corrplot)
library(ggplot2)
library(dplyr)
```

#Exploratory Data Analysis (EDA)
```{r Q1: Load and Inspect the Dataset}
data <- read_csv("kenya_student_data.csv")
#home(data)
#str(data)
#glimpse(data)
#summary(data)

#Count how many columns are numeric vs categorical
column_classes <- sapply (data, class)

# Count numeric columns 
numeric_count <- sum(column_classes == "numeric")

# Count categorical columns (often "character" or "factor")
categorical_count <- sum(column_classes == "character")

# Output the counts
cat("📊 Numeric columns:", numeric_count, "\n")
cat("🔤 Categorical columns:", categorical_count, "\n")
```

```{r Q2: Summary Stats for Numerical Variables}
data |> select(where(is.numeric)) |>  summary()

#What insights do these provide about the data?
```

```{r Q3: Bar Plot of academic_performance}
# View class distribution - class counts
table(data$academic_performance)

#Plot bargraph
data |> ggplot(aes(academic_performance)) +
        geom_bar(fill = "skyblue") +
        labs(title = "Distribution of Academic Performance",
            x = "Academic Performance",
            y = "Count") +
         theme_minimal()

#Is the target variable balanced across its classes?
#Yes, the bar heights are roughly equal
```

```{r Q4: Histogram of study_hours_weekly (Urban vs Rural)}
data |> ggplot(aes(study_hours_weekly)) +
        geom_histogram (binwidth = 2, fill = "darkorange", color = "white") +
        facet_wrap(~residency) +
        labs(title = "Study Hours by Residency", 
             x = "Study Hours per Week",
            y = "Number of Students") +
        theme_minimal()

#How does it vary between urban and rural students? DO URBAN STUDENTS STUDY MORE? 
#yes, urban students generally study more than rural students
#Look for: Shifts in the histogram center. Differences in spread or skewness. Outliers or unusual clusters.
```

```{r Q5: Boxplots of math_score by performance and gender}
data |> ggplot(aes(x = academic_performance, y = math_score)) +
        geom_boxplot(aes(fill = gender)) +
        labs (title= "Math Score by Academic Performance and Gender",
                x = "Academic Performance",
                y = "Math Score") +
        theme_minimal()

 #What patterns do you observe?       
#What to look for in the plot
#Median math score across performance categories — does it increase from "Low" to "High"?
#Spread of scores — are some categories more variable than others?
#Gender differences — do males and females follow the same patterns?
#Outliers — any extreme math scores?
```

```{r Q6: Proportions in extracurricular_activities and faculty}
# Proportions for extracurricular_activities
data |> count(extracurricular_activities) |> 
        mutate(prop = n/ sum(n) *100)

# Proportions for faculty
data |> count(faculty) |> 
        mutate(prop = n/ sum(n) *100)

#Which categories are most common?
```

```{r Q7: Correlation Matrix (exclude student_id)}
# Select the numeric variables, excluding 'student_id'
num_data <- data |>  
    select(where(is.numeric), -student_id)

# Compute correlation matrix
corr_matrix <- cor(num_data, use = "complete.obs")

# Visualize correlation matrix using a heatmap
#corrplot(corr_matrix, method = "color", type = "lower")
corrplot(corr_matrix, method = "color", type = "lower", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black",
         number.cex = 0.7)

#Which pairs have the strongest correlations?
#Look for: Strong positive correlations close to +1 (e.g., math_score vs science_score)
#Strong negative correlations close to -1 (e.g., possibly absenteeism vs academic_performance_index)
#Low correlations around 0 indicate weak or no linear relationship.
```

```{r Q8: Chi-squared Test (internet_access vs academic_performance)}
## Run chi-squared test using contingency table
chisq.test(table(data$internet_access, data$academic_performance))

#Explain p-value: significant or not? There is statistical significance
#If p < 0.05 → there is a statistically significant association between internet_access and academic_performance.
#If p ≥ 0.05 → no significant association; they are likely independent.
```
#Data Preprocessing: Missing Values
```{r Q9: Identify Missing Values}
# Calculate number and percentage of missing values per column
missing_summary <- sapply(data, function(x) sum(is.na(x)) / length(x) * 100)
missing_summary[missing_summary > 0]

#Why might these variables have missing data in a Kenyan context?
```
Why These Might Be Missing (Contextual Insight):
Family Income (5%) -Sensitive financial info; students may not know or may avoid sharing.
→ In Kenya, discussing income is often culturally sensitive, especially in lower-income or rural households.
→ Students may also not be aware of exact figures.

Attendance Rate (5%) -Could be due to poor record-keeping or informal school systems.
→ Many schools (especially rural) lack digitized records; paper-based systems often have gaps.
→ Could also reflect inconsistent attendance tracking.

Math Score (3%) -Could be ungraded students, recent transfers, or exam absences.
→ Students who missed exams due to illness, unpaid fees, or transfer cases.
→ Some schools might not submit complete results.

Academic Performance (7.82%) -May be missing if performance is yet to be evaluated or withheld.
→ Could be missing if students are newly admitted, assessments are pending, or results are withheld due to fees.

```{r Q10: Impute family_income and math_score with median}

```

