---
title: "Assignment 1- EDA and Data Preprocessing on Kenyan Student
Dataset"
output: pdf_document
date: "2025-06-12"
---
```{r Install required libraries}
install.packages(c("tidyverse","corrplot", "ggplot2","dplyr"))

#Load the libraries
library(tidyverse)
library(corrplot)
library(ggplot2)
library(dplyr)
```

#Exploratory Data Analysis (EDA)
```{r Q1: Load and Inspect the Dataset}
data <- read_csv("kenya_student_data.csv")
#home(data)
#str(data)
#glimpse(data)
#summary(data)

#Count how many columns are numeric vs categorical
column_classes <- sapply (data, class)

# Count numeric columns 
numeric_count <- sum(column_classes == "numeric")

# Count categorical columns (often "character" or "factor")
categorical_count <- sum(column_classes == "character")

# Output the counts
cat("📊 Numeric columns:", numeric_count, "\n")
cat("🔤 Categorical columns:", categorical_count, "\n")
```

```{r Q2: Summary Stats for Numerical Variables}
data |> select(where(is.numeric)) |>  summary()

#What insights do these provide about the data?
```

```{r Q3: Bar Plot of academic_performance}
# View class distribution - class counts
table(data$academic_performance)

#Plot bargraph
data |> ggplot(aes(academic_performance)) +
        geom_bar(fill = "skyblue") +
        labs(title = "Distribution of Academic Performance",
            x = "Academic Performance",
            y = "Count") +
         theme_minimal()

#Is the target variable balanced across its classes?
#Yes, the bar heights are roughly equal
```

```{r Q4: Histogram of study_hours_weekly (Urban vs Rural)}
data |> ggplot(aes(study_hours_weekly)) +
        geom_histogram (binwidth = 2, fill = "darkorange", color = "white") +
        facet_wrap(~residency) +
        labs(title = "Study Hours by Residency", 
             x = "Study Hours per Week",
            y = "Number of Students") +
        theme_minimal()

#How does it vary between urban and rural students? DO URBAN STUDENTS STUDY MORE? 
#yes, urban students generally study more than rural students
#Look for: Shifts in the histogram center. Differences in spread or skewness. Outliers or unusual clusters.
```

```{r Q5: Boxplots of math_score by performance and gender}
data |> ggplot(aes(x = academic_performance, y = math_score)) +
        geom_boxplot(aes(fill = gender)) +
        labs (title= "Math Score by Academic Performance and Gender",
                x = "Academic Performance",
                y = "Math Score") +
        theme_minimal()

 #What patterns do you observe?       
#What to look for in the plot
#Median math score across performance categories — does it increase from "Low" to "High"?
#Spread of scores — are some categories more variable than others?
#Gender differences — do males and females follow the same patterns?
#Outliers — any extreme math scores?
```

```{r Q6: Proportions in extracurricular_activities and faculty}
# Proportions for extracurricular_activities
data |> count(extracurricular_activities) |> 
        mutate(prop = n/ sum(n) *100)

# Proportions for faculty
data |> count(faculty) |> 
        mutate(prop = n/ sum(n) *100)

#Which categories are most common?
```

```{r Q7: Correlation Matrix (exclude student_id)}
# Select the numeric variables, excluding 'student_id'
num_data <- data |>  
    select(where(is.numeric), -student_id)

# Compute correlation matrix
corr_matrix <- cor(num_data, use = "complete.obs")

# Visualize correlation matrix using a heatmap
corrplot(corr_matrix, method = "color", type = "lower")

corrplot(corr_matrix, method = "color", type = "lower", 
         tl.col = "red", tl.srt = 45, addCoef.col = "black",
         number.cex = 0.7,  col = colorRampPalette(c("blue", "white", "red"))(200))

#Which pairs have the strongest correlations?
#Look for: Strong positive correlations close to +1 (e.g., math_score vs science_score)
#Strong negative correlations close to -1 (e.g., possibly absenteeism vs academic_performance_index)
#Low correlations around 0 indicate weak or no linear relationship.
```

```{r Q8: Chi-squared Test (internet_access vs academic_performance)}
# Create a contingency table
cont_table <- table(data$internet_access, data$academic_performance)

# Run chi-squared test
chisq.test(cont_table)


#Explain p-value: significant or not? There is statistical significance
#If p < 0.05 → there is a statistically significant association between internet_access and academic_performance.
#If p ≥ 0.05 → no significant association; they are likely independent.
```
#Data Preprocessing: Missing Values
```{r Q9: Identify Missing Values}
# Calculate number and percentage of missing values per column
missing_summary <- sapply(data, function(x) sum(is.na(x)))
missing_percent <- round(missing_summary / nrow(data) *100, 2)

# Combine and display
missing_report <- data.frame( Column = names(missing_summary),
                              Missing_Count = missing_summary,
                              Missing_Percent = missing_percent)

# Show only columns with missing data
missing_report[missing_report$Missing_Count > 0, ]

#Why might these variables have missing data in a Kenyan context?
```
Why These Might Be Missing (Contextual Insight):
Family Income (5%) -Sensitive financial info; students may not know or may avoid sharing.
→ In Kenya, discussing income is often culturally sensitive, especially in lower-income or rural households.
→ Students may also not be aware of exact figures.

Attendance Rate (5%) -Could be due to poor record-keeping or informal school systems.
→ Many schools (especially rural) lack digitized records; paper-based systems often have gaps.
→ Could also reflect inconsistent attendance tracking.

Math Score (3%) -Could be ungraded students, recent transfers, or exam absences.
→ Students who missed exams due to illness, unpaid fees, or transfer cases.
→ Some schools might not submit complete results.

Academic Performance (7.82%) -May be missing if performance is yet to be evaluated or withheld.
→ Could be missing if students are newly admitted, assessments are pending, or results are withheld due to fees.

```{r Q10: Impute family_income and math_score with median}
# Impute family_income with median
median_income <- median(data$family_income, na.rm = TRUE)
data$family_income[is.na(data$family_income)] <- median_income

# Impute math_score with median
median_math <- median(data$math_score, na.rm = TRUE)
data$math_score[is.na(data$math_score)] <- median_math

```

✅ Why Use the Median?
Robust to outliers:
Income and exam scores can have extreme values (e.g., very high earners or top scorers). The median is not affected by these outliers, making it more robust than the mean.

Skewed distributions:
Both family_income and math_score are often right-skewed (especially income). The median better represents the center of such distributions.

```{r Q11: Impute attendance_rate with mean + Histogram}
# Histogram before imputation
data |> ggplot(aes(attendance_rate)) +
      geom_histogram(fill = "skyblue", color = "black", na.rm = TRUE) +
      labs(title = "Attendance Rate Before Imputation", x = "Attendance Rate", y = "Count")

# Impute missing values with the mean
data$attendance_rate[is.na(data$attendance_rate)] <- mean(data$attendance_rate, na.rm = TRUE)

# Histogram after imputation
data |> ggplot(aes(attendance_rate)) +
        geom_histogram(fill = "tomato", color = "black") +
        labs(title = "Attendance Rate After Imputation", x = "Attendance Rate", y = "Count")

```
📝 Why Use the Mean Here?
attendance_rate is a continuous variable.

The distribution is typically symmetric or mildly skewed for attendance.

Mean imputation retains the central tendency and is acceptable when the missingness is random and small in proportion (5% in your case).

```{r Compare the distributions before and after imputation using histograms.}
# Copy of original attendance_rate before imputation
data$attendance_rate_original <- data$attendance_rate

#Impute missing values using the mean
data$attendance_rate[is.na(data$attendance_rate)] <- mean(data$attendance_rate, na.rm = TRUE)

library(dplyr)
library(tidyr)

# Prepare data for plotting
attendance_df <- data |> 
    select(attendance_rate_original, attendance_rate) |> 
    pivot_longer(cols = everything(), names_to = "Version", values_to = "Rate") |> 
    mutate(Version = recode(Version,
                          "attendance_rate_original" = "Before Imputation",
                          "attendance_rate" = "After Imputation"))

# Plot overlaid histograms
ggplot(attendance_df, aes(x = Rate, fill = Version)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(title = "Overlayed Histograms of Attendance Rate",
       x = "Attendance Rate", y = "Count") +
  scale_fill_manual(values = c("steelblue", "orange")) +
  theme_minimal()


```
Statistical Impact
Visual inspection will show a small peak at the mean value in the "After Imputation" version due to imputed entries.

Effect on distribution is minor if missingness is small (~5%), but the tail behavior may be slightly compressed.

Since the mean preserves overall central tendency, it's acceptable when data is fairly symmetric — if highly skewed, median is safer.

#Data Preprocessing: Outliers
```{r Q12: Detect Outliers (IQR method)}
# Compute Q1, Q3, and IQR
Q1 <- quantile(data$family_income, 0.25, na.rm = TRUE)
Q3 <- quantile(data$family_income, 0.75, na.rm = TRUE)
IQR<- Q3 - Q1

# Define outlier thresholds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Detect outliers
outliers <- length(data$family_income[data$family_income < lower_bound | data$family_income > upper_bound])

# Print results
cat("Number of outliers in family income:", outliers, "\n")

#what might they represent in a Kenyan context?
```
🧠 Interpretation of outliers in family_income may represent:

High-income families in affluent areas (e.g., Nairobi suburbs, expatriate households, upper-class).

Low outliers may reflect students from marginalized or rural communities.

Could indicate data entry errors (e.g., extra zero added or wrong currency).

Socioeconomic inequality is pronounced in Kenya, so such income variation is realistic and informative.

```{r Q13: Cap Outliers + Boxplot}
# Cap values 
data$family_income_capped <- data$family_income
data$family_income_capped[data$family_income > upper_bound] <- upper_bound
data$family_income_capped[data$family_income < lower_bound] <- lower_bound

library(tidyr)

# Prepare data for plotting
income_df <- data |> 
  select(family_income, family_income_capped) |> 
  pivot_longer(cols = everything(), names_to = "Version", values_to = "Income") |> 
  mutate(Version = recode(Version,
                          "family_income" = "Before Capping",
                          "family_income_capped" = "After Capping"))

# Plot boxplots
ggplot(income_df, aes(x = Version, y = Income, fill = Version)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1, width = 0.6) +
  labs(title = "Boxplot of Family Income Before and After Capping",
       x = "", y = "Family Income") +
  theme_minimal()

```
🧠 Interpretation
Before Capping: You’ll see extreme values (outliers) far from the box.

After Capping: Outliers are pulled within the upper/lower bounds; reduces skewness and extreme influence.

This is useful for robust statistical modeling.

Capping outliers (also called winsorizing) is a data preprocessing technique where extreme values (outliers) are replaced with a threshold value instead of being removed. This helps reduce the influence of outliers without losing data points entirely.

🔍 Why cap outliers?
Outliers can:

Skew statistical summaries (like mean, standard deviation)

Distort visualizations (like histograms, boxplots)

Affect machine learning model performance

✅ When to use capping
When removing outliers would lose valuable data

When models are sensitive to extreme values (like linear regression)

#Data Preprocessing: Feature Engineering
```{r Q14.Discretize study_hours_weekly into bins}
# use the cut() function to categorize the continuous variable; discretize into 4 bins ***
data$study_hours_categorized <- cut(data$study_hours_weekly,
                      breaks = c(-Inf, 5, 10, 15, Inf), # Define fixed breakpoints
                      labels = c("Low", "Moderate", "High", "Very High"), right = TRUE)

# Bar plot of the binned variable
data |> ggplot(aes(study_hours_categorized)) +
        geom_bar(fill = "steelblue") +
        #facet_wrap(~academic_performance) +
        labs(title = "Study Hours per Week (Binned)",
             x = "Study Hours Category",
            y = "Number of Students") +
       theme_minimal()

```

```{r Q15. Discretize family_income into quartiles}
# Discretize into quartiles with meaningful labels
data <- data |> 
  mutate(income_quartile = cut(family_income,
    breaks = quantile(family_income, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE),
    include.lowest = TRUE,
    labels = c("Low", "Medium-Low", "Medium-High", "High")))

# correlation with academic_performance using contingency table (counts)
cont_table <- table(data$income_quartile, data$academic_performance)
cont_table

# Chi-squared test to statistically assess the relationship
chisq.test(cont_table)



```
🔍 Interpretation:
High income quartile has a greater proportion of students with Excellent academic performance while the low income quartile has a great proportion of students with Average academic performance. It may suggest a positive association between income and academic performance.
❌ Conclusion:
But since p > 0.05, there is no statistically significant association between family income level and academic performance.
🔍 Possible Explanations (Kenyan context):
Some students from lower-income households may receive scholarships or support that helps them perform equally well.

```{r Q16. Create average score}
# Create total_score by averaging the three subject scores
data$total_score <- rowMeans(data[, c("math_score", "science_score", "english_score")], na.rm = TRUE)

# Plot the distribution of total_score
data |> ggplot(aes(x = total_score)) +
        geom_histogram(binwidth = 4, fill = "salmon", color = "black") +
      labs(title = "Distribution of Total Score",
       x = "Total Score (Average of 3 Subjects)",
       y = "Number of Students") +
      theme_minimal()
```

#Data Preprocessing: Relationships
```{r Q17. Contingency table}
activity_perf <- table(data$extracurricular_activities, data$academic_performance)
activity_perf

#What patterns suggest about student involvement?

chisq.test(activity_perf)
```
🔍 Quick Observations
"None" group has slightly more students with Poor and Good performance compared to other categories. 
"Both" has higher counts in "Average" followed by "Poor"
"SPorts" group has the highest number of students with "Excellent" performance
—potentially suggesting a positive link between extracurricular activity and academic performance.

📌 Chisquare Interpretation
Since the p-value is much greater than 0.05, we fail to reject the null hypothesis.

This means there is no statistically significant association between extracurricular_activities and academic_performance based on your data.

🧠 What This Suggests:
Participation in extracurriculars does not have a strong or significant influence on academic performance in this dataset. There might be other confounding factors (e.g., socio-economic status, time management skills) that play a larger role.

```{r Q18: Scatter Plot}
#scatter plot between study_hours_weekly and total_score colored by residency
data |> ggplot(aes(study_hours_weekly, total_score, color = residency)) +
        geom_point(alpha = 0.6) +
        geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
        labs(title = "Study Hours vs Total Score by Residency",
              x = "Weekly Study Hours",
              y = "Total Score") +
        theme_minimal()



```
The regression line (from lm(total_score ~ study_hours_weekly)) cuts through the middle because it’s a single global trend across all students, regardless of residency.

Urban and Rural points appear similarly spread – this suggests that:
The relationship between study_hours_weekly and total_score is not strongly different between Urban and Rural students.
Or there might be very little overall correlation between study hours and total score, regardless of residency.

```{r Save preprocessed dataset}
# fitting and plotting separate regression lines
plot(data$study_hours_weekly, data$total_score,
     col = ifelse(data$residency == "Urban", "salmon", "seagreen"),
     pch = 19,
     xlab = "Weekly Study Hours",
     ylab = "Total Score",
     main = "Study Hours vs Total Score by Residency")

# Separate linear models
lm_urban <- lm(total_score ~ study_hours_weekly, data = subset(data, residency == "Urban"))
lm_rural <- lm(total_score ~ study_hours_weekly, data = subset(data, residency == "Rural"))

# Add separate trend lines
abline(lm_urban, col = "salmon", lwd = 2)
abline(lm_rural, col = "seagreen", lwd = 2)

# Add legend
legend("bottomright", legend = c("Urban", "Rural"), col = c("salmon", "seagreen"), lwd = 2)

write_csv(data, "kenya_student_data_preprocessed.csv")

```
both lines are flat → no strong relationship between study time and total score.
